\textbf{COMPARISON OF \\ HEURISTIC ALGORITHMS \\ FOR KP01} \\[0.5cm] \rule{12cm}{0.3mm} \\[0.5cm] \small \scshape{A comparison of heuristic algorithms for solving the 0-1 Knapsack Problem} \\[0.4cm] \rule{12cm}{0.3mm} \vskip -0.3cm

### 2023-04-15 doc/paper.tex
some formatting

\subsection{Solving the 0-1 Knapsack Problem}

@@

\subsubsection*{Deterministic algorithms}
Dynamic programming is a general technique for solving optimization problems. If a problem has optimal substructure and over-lapping subproblems, then dynamic programming is applicable. In computer science 
a problem has \emph{optimal substructure} if an optimal solution for a problems can be constructed from optimal solutions of its sub-problems, and \emph{over-lapping subproblems} is when a problem can be decomposed into sub-problems which are reused. Dynamic programming breaks down a complicate problem into smaller sub-problems in a recursive manner, while also using some memory to save the solutions of the sub-problems (either in tabular form or memoized form). This way, when we need to get a solution for a sub-problem again, we can just use our previously calculated value.

@@

\begin{breakablealgorithm}
\caption{Solving 0-1 Knapsack with Dynamic Programming}\label{dp}
    \begin{algorithmic}[1]
        \For {$i = 0 \text{ to } noItems $} \Comment{\textcolor{blue}{If no items, then profit = 0}}
            \State $Table[i][0] \gets 0 $ 
        \EndFor
        \For {$k = 0 \text{ to } maxCapacity $} \Comment{\textcolor{blue}{If no capacity, then profit = 0}}
            \State $Table[0][k] \gets 0 $
        \EndFor
        \For {$i = 0 \text{ to } noItems$}
            \For {$k = 0 \text{ to } maxCapacity $}
                \State $Table[i][k] \gets Table[i-1][k]$  
                \If {$k-weights[i] >= 0$} 
                    \State $Table[i][k] \gets max(Table[i-1][k-weights[i]] + profits[i], Table[i][k]) $
                \EndIf
            \EndFor
        \EndFor
        \State $\textbf{Print } Table[noItems][maxCapacity] $
    \end{algorithmic}
\end{breakablealgorithm}

@@

\subsubsection*{Heuristic algorithms}
When deterministic algorithms prove to be too slow for practical applications, heuristic algorithms are chosen for their (usually) near optimal outputs and their speed. Heuristic algorithms should not be confused with approximation algorithms. Approximation algorithms guarantee a maximum margin of error; a constant factor off of the optimal. On the other hand, a heuristic algorithm does not guarantee anything, so it can perform better or worse than an approximation algorithm.

Heuristic algorithms for combinatorial optimization are usually designed in a specific way:
\begin{enumerate}
    \item Create a potential solution which is current best.
    \item Generate new solution. (usually in polynomial time) 
    \item If new solution better than current best, swap out the best with the newly generated. 
    \item If satisfied, output best solution, else continue.
\end{enumerate}

@@

\paragraph*{Discrete Global-Best Harmony Search Algorithm}
The Harmony Search algorithm (HS) was developed in 2001 \ref{geem et al} based on the improvisation of music players \textcolor{red}{EXPAND}. This algorithm was made for continuous search spaces and thus cannot be used for discrete search spaces in combinatorial optimization problems. The Discrete Global-Best harmony search was proposed by \textcolor{red}{EXPAND} to overcome this. The DGHS for KP01 also has a repair-operator and a greedy selection mechanism. A harmony in this algorithm refers to a candidate solution, or a certain selection of items to be put in the knapsack.

@@

%DGHS
\begin{breakablealgorithm}
\caption{The DGHS algorithm}\label{DGHS}
    \begin{algorithmic}[1]
        \State Set the harmony memory size $HMS$, the number of iterations $ITERATIONS$, and the minimum and maximum values of parameters $PAR$ and $HMCR$.
        \State Initialize the HM through a randomized process, and use \ref{harmonyRepair} to the generated harmonies. Calculate the totalProfit and totalWeight values for each harmony in HM.
        \State $iterator \gets 1$
        \While {$iterator <= ITERATIONS$}
            \State Record indices of the best and the worst harmonies in HM.
            \State Calculate parameters HMCR and PAR for the current iteration.
            \State Perform Algorithm \ref{harmonyGen} to produce a new harmony $\Vec{x}_{new}$
            \State Perform Algorithm \ref{harmonyRepair} to repair the new harmony $\Vec{x}_{new}$
            \If {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{best}$}
                \State Replace $\Vec{x}_{best} \text{ with } \Vec{x}_{new}$
            \ElsIf {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{worst}$}
                \State Replace $\Vec{x}_{worst} \text{ with } \Vec{x}_{new}$
            \EndIf
            \State $iterator \gets iterator+1$
        \EndWhile
        \State Output profit of best harmony
    \end{algorithmic}
\end{breakablealgorithm}

@@

The dynamic updating of the parameters is designed in this way to let larger values of HMCR help accelerate the convergence of the harmonies early on with the the help of the best individual harmony, while smaller values of HMCR can help you overcome local maxima. Similarly, larger values of PAR in the beginning allow for increases in diversity through mutations (item decision bit flips), when there is time to "explore" different variants, smaller values allow convergence at the end of the search.

@@

%harmonyGen
\begin{breakablealgorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}[1]
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{// Flipping i'th item - mutation}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{breakablealgorithm}

@@

%harmonyRepair
\begin{breakablealgorithm}
\caption{Repair-operator for DGHS}\label{harmonyRepair}
    \begin{algorithmic}[1]
        \If {$totalWeight > maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ DROP phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
            \State Remove the ones with the least profit-density values greedily
                \If {$\lambda_{i} == 0$}
                    \State \textbf{Continue}
                \EndIf
                \State $newHarmony[ind_{i}] = 0$ \Comment{Unload the item}
                \State $totalWeight \gets totalWeight - weight[ind_{i}]$
                \State $totalProfit \gets totalProfit - profit[ind_{i}]$
                \If {$totalWeight <= maxWeight$}
                    \State \textbf{Break} \Comment{Terminate DROP phase}
                \EndIf
            \EndFor
        \EndIf
        \If {$totalWeight < maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ADD phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
                \State Add the ones with the greatest profit-density if possible 
                \If {$newHarmony[ind_{i}] == 0$}
                    \If {$totalWeight + weight[ind_{i}] <= maxWeight$}
                        \State $newHarmony[ind_{i}] = 1$
                        \State $totalWeight \gets totalWeight + weight[ind_{i}]$
                        \State $totalProfit \gets totalProfit + profit[ind_{i}]$
                    \EndIf
                \EndIf
            \EndFor
        \EndIf
    \end{algorithmic}
\end{breakablealgorithm}

\paragraph*{Binary Multi-Scale Quantum Harmonic Oscillator Algorithm}

%BMQHOA
\begin{breakablealgorithm}
\caption{The BMQHOA algorithm with solution generation}\label{BMQHOA}
    \begin{algorithmic}[1]
        \State Set the number of iterations $ITERATIONS$ and the number of binary vectors (solutions) - BINVEC in memory.
        \State Randomly generate the binary vectors and use Algorithm \ref{harmonicRepair} to repair the vectors. 
        \While {$iterator <= ITERATIONS$}
            \State Update $\sigma_{s}$
            \State $found \gets FALSE$
            \While {$found == FALSE$}
                \State Try to generate a solution
                \State Let $solutions_{new}$ be the new generated vector 
                \State $solutions_{new} \gets solutions_{best}$
                \State Generate the number of flipped bits $m ~ N(0, \sigma_{s})$
                \State Treat $solutions_{new}$ as a circular array
                \State Randomly select a position in $solutions_{new}$ and flip the next $m$ items.
                \State Mutate a random bit towards current best solution (flip an item)
                \State $solutions_{new}[rand] \gets solutions_{best}[rand]$
                \State Repair newly generated solution
                \If {$totalProfit >= totalProfit_{worst}$}
                    \State Replace worst solution with newly generated solution
                    \State $solutions_{worst} \gets solutions_{new}$
                    \State $found = TRUE$
                \EndIf
            \EndWhile
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}

After generating a new solution, $m$ bits are flipped in the hope of increasing diversity in the solutions and increasing the likelihood of finding the optimal solution. They are generated randomly with a normal probability distribution around 0, with a variable std deviation. The Std. Deviation decreases with each new-solution-generation iteration so as to increase diversity in the beginning of the search, while reducing the likelihood of corrupting a binary vector closer to the end of the search when close to optimal solutions have (hopefully) been found. The normal distribution probability density function generator is implemented with the C++ Standard Library class normal\_distribution. An item's decision value is also made to match the corresponding decision in the best solution, which gives a slow mutation towards the current best solution allowing for diversity while still allowing the best solution to influence the process. Allowing multiple bits to mutate toward current best, this can lead to a premature local maximum, nullifying the rest of the search \ref{BMQHOA-article} .

@@

\begin{breakablealgorithm}
\caption{Repair-Operator for BMQHOA} 
    \begin{algorithmic}
        \State Let $x$ be the current array/vector.
        \State $totalWeight = 0, temp = 0, i = 0, b = 0$
        \State Stage 1: Density first stage
        \While {$temp < maxWeight$}
            \State $totalWeight = temp$
            \State $i += 1$
            \State $temp = temp + weight[Q_{1}[i]]$
        \EndWhile
        \For {$j = i \text{ to } n$}
            \State $x[Q_{1}[j]] = 0$
            \State $totalWeight += x[Q_{1}[j]]$
        \EndFor
        \State Stage 2: Minimum weight first stage
        \While {$totalWeight < C$}
            \State $b += 1$
            \If {$totalWeight + weight[Q_{2}[b]] <= C$} 
                    \State $x[Q_{2}[b]] = 1$
                \State $totalWeight += weight[Q_{2}[b]]$
            \EndIf
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}

>>>>

\paragraph{Solving the 0-1 Knapsack Problem}\mbox{}\

@@

\subsection*{Deterministic algorithms}
Dynamic programming is a general technique for solving optimization problems. If a problem has optimal substructure and over-lapping subproblems, then dynamic programming is applicable. In computer science 
a problem has \emph{optimal substructure} if an optimal solution for a problems can be constructed from optimal solutions of its sub-problems, and \emph{over-lapping subproblems} is when a problem can be decomposed into sub-problems which are reused. Dynamic programming breaks down a complicate problem into smaller sub-problems in a recursive manner, while also using some memory to save the solutions of the sub-problems (usually in tabular form). This way, when we need to get a solution for a sub-problem again, we can just use our previously calculated value.

@@

%dp
\begin{breakablealgorithm}
\caption{Solving 0-1 Knapsack with Dynamic Programming}\label{dp}
    \begin{algorithmic}[1]
        \For {$i = 0 \text{ to } noItems $} \Comment{\textcolor{blue}{If no items, then profit = 0}}
            \State $Table[i][0] \gets 0 $ 
        \EndFor
        \For {$k = 0 \text{ to } maxCapacity $} \Comment{\textcolor{blue}{If no capacity, then profit = 0}}
            \State $Table[0][k] \gets 0 $
        \EndFor
        \For {$i = 0 \text{ to } noItems$}
            \For {$k = 0 \text{ to } maxCapacity $}
                \State $Table[i][k] \gets Table[i-1][k]$  
                \If {$k-weights[i] >= 0$} 
                    \State $Table[i][k] \gets max(Table[i-1][k-weights[i]] + profits[i], Table[i][k]) $
                \EndIf
            \EndFor
        \EndFor
        \State $\textbf{Print } Table[noItems][maxCapacity] $
    \end{algorithmic}
\end{breakablealgorithm}

\vskip 0.5cm

@@

\subsection*{Heuristic algorithms}
When deterministic algorithms prove to be too slow for practical applications, heuristic algorithms are chosen for their (usually) near optimal outputs and their speed. Heuristic algorithms should not be confused with approximation algorithms. Approximation algorithms guarantee a maximum margin of error; a constant factor off of the optimal. On the other hand, a heuristic algorithm does not guarantee anything, so it can perform better or worse than an approximation algorithm.

Heuristic algorithms for combinatorial optimization are usually designed in a specific way \ref{some website}:
\begin{enumerate}
    \item Create a potential solution which is current best.
    \item Generate new solution. (usually in polynomial time) 
    \item If new solution better than current best, swap out the best with the newly generated. 
    \item If satisfied, output best solution, else continue.
\end{enumerate}

@@

\subsubsection{Discrete Global-Best Harmony Search Algorithm}
The Harmony Search algorithm (HS) was developed in 2001 \ref{geem et al} based on the improvisation of music players \textcolor{red}{EXPAND}. This algorithm was made for continuous search spaces and thus cannot be used for discrete search spaces in combinatorial optimization problems. The Discrete Global-Best harmony search was proposed by \textcolor{red}{EXPAND} to overcome this. The DGHS for KP01 also has a repair-operator and a greedy selection mechanism. A harmony in this algorithm refers to a candidate solution, or a certain selection of items to be put in the knapsack.

@@

\vskip 0.5cm
%DGHS
\begin{breakablealgorithm}
\caption{The DGHS algorithm}\label{DGHS}
    \begin{algorithmic}[1]
        \State Set the harmony memory size $HMS$, the number of iterations $ITERATIONS$, and the minimum and maximum values of parameters $PAR$ and $HMCR$.
        \State Initialize the HM through a randomized process, and use \ref{harmonyRepair} to the generated harmonies. Calculate the totalProfit and totalWeight values for each harmony in HM.
        \State $iterator \gets 1$
        \While {$iterator <= ITERATIONS$}
            \State Record indices of the best and the worst harmonies in HM.
            \State Calculate parameters HMCR and PAR for the current iteration.
            \State Perform Algorithm \ref{harmonyGen} to produce a new harmony $\Vec{x}_{new}$
            \State Perform Algorithm \ref{harmonyRepair} to repair the new harmony $\Vec{x}_{new}$
            \If {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{best}$}
                \State Replace $\Vec{x}_{best} \text{ with } \Vec{x}_{new}$
            \ElsIf {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{worst}$}
                \State Replace $\Vec{x}_{worst} \text{ with } \Vec{x}_{new}$
            \EndIf
            \State $iterator \gets iterator+1$
        \EndWhile
        \State Output profit of best harmony
    \end{algorithmic}
\end{breakablealgorithm}
\vskip 0.5cm

@@

The dynamic updating of the parameters is designed in this way to let larger values of HMCR help accelerate the convergence of the harmonies early on with the the help of the best individual harmony, while smaller values of HMCR can help you overcome local maxima \ref{DGHS-algorithm}. Similarly, larger values of PAR in the beginning allow for increases in diversity through mutations (item decision bit flips), when there is time to "explore" different variants, smaller values allow convergence at the end of the search.

@@

%harmonyGen
\begin{breakablealgorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}[1]
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{\textcolor{blue}{Flipping i'th item - mutation}}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{breakablealgorithm}
\vskip 0.5cm

@@

\vskip 0.5cm
%harmonyRepair
\begin{breakablealgorithm}
\caption{Repair-operator for DGHS}\label{harmonyRepair}
    \begin{algorithmic}[1]
        \If {$totalWeight > maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{\textcolor{blue}{DROP phase}}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
            \State Remove the ones with the least profit-density values greedily
                \If {$\lambda_{i} == 0$}
                    \State \textbf{Continue}
                \EndIf
                \State $newHarmony[ind_{i}] = 0$ \Comment{\textcolor{blue}{Unload the item}}
                \State $totalWeight \gets totalWeight - weight[ind_{i}]$
                \State $totalProfit \gets totalProfit - profit[ind_{i}]$
                \If {$totalWeight <= maxWeight$}
                    \State \textbf{Break} \Comment{\textcolor{blue}{Terminate DROP phase}}
                \EndIf
            \EndFor
        \EndIf
        \If {$totalWeight < maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{\textcolor{blue}{ADD phase}}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
                \State Add the ones with the greatest profit-density if possible 
                \If {$newHarmony[ind_{i}] == 0$}
                    \If {$totalWeight + weight[ind_{i}] <= maxWeight$}
                        \State $newHarmony[ind_{i}] = 1$
                        \State $totalWeight \gets totalWeight + weight[ind_{i}]$
                        \State $totalProfit \gets totalProfit + profit[ind_{i}]$
                    \EndIf
                \EndIf
            \EndFor
        \EndIf
    \end{algorithmic}
\end{breakablealgorithm}

\vskip 1cm
\subsubsection{Binary Multi-Scale Quantum Harmonic Oscillator Algorithm}

The Multi-Scale Quantum Harmonic Oscillator Algorithm (MQHOA) is called such as it follows a model of a solving a particle's ground state wave function under the harmonic oscillator potential well \ref{BMQHOA-article} . In MQHOA, candidate solutions are generated by sampling points in a Gaussian distribution within a certain distance of something. The Binary Multi-Scale Quantum Harmonic Oscillator Algorithm (BMQHOA) discretizes this by defining the number of bits between solutions (item decision variable ) as the distance between solutions, so it becomes a discrete search space. Similar to the DGHS algorithm, a repair-operator is added fix solutions which violate the capacity constraint.

The BMQHOA algorithm consists of 4 parts:
\begin{enumerate}
    \item Random binary vector generation -> repair
    \item Generating a solutions by flipping $m$ items, mutating, and repairing
    \item Reducing the standard deviation value for the normal distribution from which $m$ is determined every iteration.
    \item If termination constraint is satisfied, output best totalProfit value, else return to Step 2.
\end{enumerate}

%BMQHOA
\begin{breakablealgorithm}
\caption{The BMQHOA algorithm with solution generation}\label{BMQHOA}
    \begin{algorithmic}[1]
        \State Set the number of iterations $ITERATIONS$ and the number of binary vectors (solutions) - BINVEC in memory.
        \State Randomly generate the binary vectors and use Algorithm \ref{harmonicRepair} to repair the vectors. 
        \While {$iterator <= ITERATIONS$}
            \State Update $\sigma_{s}$
            \State $found \gets FALSE$
            \While {$found == FALSE$}
                \State Try to generate a solution
                \State Let $solutions_{new}$ be the new generated vector 
                \State $solutions_{new} \gets solutions_{best}$
                \State Generate the number of flipped bits $m ~ N(0, \sigma_{s})$
                \State Treat $solutions_{new}$ as a circular array
                \State Randomly select a position in $solutions_{new}$ and flip the next $m$ items.
                \State Mutate a random bit towards current best solution (flip an item)
                \State $solutions_{new}[rand] \gets solutions_{best}[rand]$
                \State Repair newly generated solution
                \If {$totalProfit >= totalProfit_{worst}$}
                    \State Replace worst solution with newly generated solution
                    \State $solutions_{worst} \gets solutions_{new}$
                    \State $found = TRUE$
                \EndIf
            \EndWhile
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}
\vskip 0.5cm

After generating a new solution, $m$ items are flipped in the hope of increasing diversity in the solutions and increasing the likelihood of finding the optimal solution. They are generated randomly with a normal probability distribution around \textbf{0}, with a mean of $0$ and a variable std deviation. The Std. Deviation decreases with each new-solution-generation iteration so as to increase diversity in the beginning of the search, while reducing the likelihood of corrupting a binary vector closer to the end of the search when close to optimal solutions have (hopefully) been found. Initially the value of the standard deviation is set to $noItems/3$, so that there is a $\sim 99.7\%$ chance that the generated number of flipped items $m$ are in the range $[0-3\sigma_{s}, 0+3\sigma_{s}]$:
$$STDEV_{max} = noItems/3$$
$$\sigma_{s} = 1-\frac{t}{ITERATIONS} STDEV_{max}$$

The normal distribution probability density function generator is implemented with the C++ Standard Library class normal\_distribution. An item's decision value is also made to match the corresponding decision in the best solution, which gives a slow mutation towards the current best solution allowing for diversity while still allowing the best solution to influence the process. Allowing multiple bits to mutate toward current best, this can lead to a premature local maximum, nullifying the rest of the search \ref{BMQHOA-article} .

@@

\begin{breakablealgorithm}
\caption{Repair-Operator for BMQHOA}\label{harmonicRepair}
    \begin{algorithmic}
        \State Let $x$ be the current array/vector.
        \State $totalWeight = 0, temp = 0, i = 0, b = 0$
        \State Stage 1: Density first stage
        \While {$temp < maxWeight$}
            \State $totalWeight = temp$
            \State $i += 1$
            \State $temp = temp + weight[Q_{1}[i]]$
        \EndWhile
        \For {$j = i \text{ to } n$}
            \State $x[Q_{1}[j]] = 0$
            \State $totalWeight += x[Q_{1}[j]]$
        \EndFor
        \State Stage 2: Minimum weight first stage
        \While {$totalWeight < C$}
            \State $b += 1$
            \If {$totalWeight + weight[Q_{2}[b]] <= C$} 
                    \State $x[Q_{2}[b]] = 1$
                \State $totalWeight += weight[Q_{2}[b]]$
            \EndIf
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}
